An optimization of bidirectional update language based on the idea of reversible computing \thanks{Supported by organization x.}

### 2019-09-13 paper/main.tex
update tex 20190913-7.53
update from minbigul -> experiment

\newcommand{\sif}[3]{\text{if } #1 \text{ then } #2 \text{ else } #3}
\newcommand{\product}[2]{#1 \ prod \ #2}
\newcommand{\tuple}[2]{(#1 :: #2)}
\newcommand{\rearrs}[3]{RearrS \ #1 \ #2 \ #3}
\newcommand{\rearrv}[3]{RearrV \ #1 \ #2 \ #3}
\newcommand{\casebx}[4]{Case \ #1 \ #2 \ #3 \ #4}

@@

\begin{itemize}
\item our target language is minBiGUL (BiGUL minus adaptive case) : very well-behaved
\item Basic explanation of bidirectional transformation
  \begin{itemize}
  \item get = s $\to$ v
  \item put = s * v $\to$ s
  \item explanation about get and put.
  \item We assume putget and getput hold (explanation about putget and getput)
  \item Because our target language satisfies very well-behaved (No adaptive case) $\to$ putput hold.
  \end{itemize}
\item explanation about put-based BX language: BiGUL
\end{itemize}

@@

$data := Unit \\ 
    \qtab | \ Int \ x \\ 
    \qtab | \ Con(data_1, data_2)$\\

$bx := Skip \ h \\ 
    \qtab | \ Replace \\ 
    \qtab | \ bx_1 \ prod \ bx_2 \ \\ 
    \qtab | \ RearrS \ f_1 \ f_2 \ bx \\
    \qtab | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtab | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \\ 
    \qtab | \ bx_1 \circ bx_2$

\subsection{Semantics of minBiGUL}

@@

$\putbx{\product{bx_1}{bx_2}}{(s_1 :: s_2)}{(v_1 :: v_2)} =\\
    \tab (\putbx{bx_1}{s_1}{v_1}) :: (\putbx{bx_2}{s_2}{v_2})$

@@

$\putbx{bx_1 \circ bx_2}{s}{v} = \putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}$

@@

$\getbx{\product{bx_1}{bx_2}}{(s_1 :: s_2)} = \\
    \tab (\getbx{bx_1}{s_1}) :: (\getbx{bx_2}{s_2})$

@@

$\getbx{bx_1 \circ bx_2}{s} = \getbx{bx_2}{(\getbx{bx_1}{s})}$

\begin{itemize}
\item To implement these function in OCaml, we need some way to produce inverse function for $f$ and $g$.
\item $f_2 = f_1^{-1}, g_2 = g_1^{-1}$
\end{itemize}

@@

For the approaches so far, when evaluating a composition of BX programs, $get$s are re-evaluated so many times since no intermediate state is stored during the evaluation. This is a kind of information loss. One question is whether it is possible to calculate the above programs without losing information. And the answer is yes. It comes from the the idea of reversible computation where all information during the evaluation need to be kept. In minBiGUL, we can do that by tupling $put$ and $get$. The such pair is accepted as input of a function named $pg$ to produce a new one that contains the actual result of a BX program.

@@

$\pg{\product{bx_1}{bx_2}}{s_1 :: s_2}{v_1 :: v_2} \\
    \tab = ((\putbx{bx_1}{s_1}{v_1}) :: (\putbx{bx_2}{s_2}{v_2}), (\getbx{bx_1}{s_1}) :: (\getbx{bx_2}{s_2})) \\
    \tab = (s_1, v_1) \Leftarrow \pg{bx_1}{s_1}{v_1}; \\
        \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{s_2}{v_2}; \\
        \qtab \tab (s_1 :: s_2, v_1 :: v_2)$

@@

Constructions of $pg$ for $[\![Replace]\!]$, $[\![\product{bx_1}{bx_2}]\!]$, $[\![\rearrs{f_1}{f_2}{bx}]\!]$ and $[\![\rearrv{g_1}{g_2}{bx}]\!]$ are very clear when just paring $put$ and $get$ respectively, then doing basic changes.\\

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling $put$ and $get$ occurs, there will be 4 combinations of these conditions. This means that two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satified. The evaluated BXs are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others, which the construction of $\pginline{Case}$ uses above. \\

@@

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. It is possible if we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that make the $put \, [\![bx_1]\!]$ valid. The last equality substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which is the evaluated result of $\putbx{bx_1}{s}{dummy}$. This transformation is feasible due to the PUTPUT law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

To better understand about $pg$, let's take a look at some examples. The following is the definition of $phead$ in BiGUL [?]:

    $\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where:}\\
        \qtab f_1 = \lambda (s:ss).(s,ss)\\
        \qtab f_2 = \lambda (s,ss).(s:ss)\\
        \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where:}\\
            \qtab \qtab g_1 = \lambda v.(v,())\\
            \qtab \qtab g_2 = \lambda (v,()).v\\
            \qtab \qtab bx_v = Replace \ prod \ (Skip (const ()))$\\

Now, we evaluate $\putbxinline{phead}$ and $\getbxinline{phead}$ based on their definitions:

$\putbx{phead}{[1,2,3]}{100}\\
    \tab = \putbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}{100}\\
    \tab = f_2 \ (\putbx{bx_s}{(f_1 \ [1,2,3])}{100})\\
    \tab = f_2 \ (\putbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}{100})\\
    \tab = f_2 \ (\putbx{bx_v}{(1,[2,3])}{(g_1 \ 100}))\\
    \tab = f_2 \ (\putbx{Replace \ prod \ (Skip (const ()))}{(1,[2,3])}{(100,())})\\
    \tab = f_2 \ ((\putbx{Replace}{1}{100}) :: (\putbx{(Skip (const ()))}{[2,3]}{()}))\\
    \tab = f_2 \ (100 :: [2,3])\\
    \tab = f_2 \ (100, [2,3])\\
    \tab = [100,2,3]$\\

$\getbx{phead}{[1,2,3]}\\
    \tab = \getbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}\\
    \tab = \getbx{bx_s}{(f_1 \ [1,2,3])}\\
    \tab = \getbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}\\
    \tab = g_2 \ (\getbx{bx_v}{(1,[2,3])})\\
    \tab = g_2 \ (\getbx{Replace \ prod \ (Skip (const ()))}{(1,[2,3])})\\
    \tab = g_2 \ ((\getbx{Replace}{1}) :: (\getbx{(Skip (const ()))}{[2,3]}))\\
    \tab = g_2 \ (1 :: ())\\
    \tab = g_2 \ (1,())\\
    \tab = 1$\\

@@

On the other hand, we can evaluate $\pginline{phead}$ due to the construction of $pg$:
$\pg{phead}{[1,2,3]}{100}\\
    \tab = \pg{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}{100}\\
    \tab = (s, v) \Leftarrow \pg{bx_s}{f_1 \ [1,2,3]}{100};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \pg{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}{100};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \pg{bx_v}{(1,[2,3])}{g_1 \ 100};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \pg{Replace \ prod \ (Skip (const ()))}{(1,[2,3])}{((100,()))};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \{\\
            \qtab \qtab (s_1, v_1) \Leftarrow \pg{Replace}{1}{100};\\
            \qtab \qtab (s_2, v_2) \Leftarrow \pg{(Skip (const ()))}{[2,3]}{()};\\
            \qtab \qtab (s_1 :: s_2, v_1 :: v_2)\\
        \qtab \tab \};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \{\\
            \qtab \qtab (s_1, v_1) \Leftarrow (100, 1);\\
            \qtab \qtab (s_2, v_2) \Leftarrow ([2,3], ());\\
            \qtab \qtab (s_1 :: s_2, v_1 :: v_2)\\
        \qtab \tab \};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow ((100,[2,3]), (1,()))\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow ((100, [2,3]), 1);\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = ([100,2,3], 1)$\\

    \tab $\putbx{phead \circ phead}{[[1,2,3],[\ ],[4,5]]}{100} = [[100,2,3],[\ ],[4,5]]$\\
    \tab $\getbx{phead \circ phead}{[[1,2,3],[\ ],[4,5]]} = 1$\\

$\getbx{phead_1 \circ phead_2}{[[1,2,3],[\ ],[4,5]]}\\
    \tab = \getbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\ ],[4,5]]})}\\
    \tab = \getbx{phead_2}{[1,2,3]}\\
    \tab = 1$\\

$\putbx{phead_1 \circ phead_2}{[[1,2,3],[\ ],[4,5]]}{100}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\ ],[4,5]]}{(\putbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\ ],[4,5]]})}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\ ],[4,5]]}{(\putbx{phead_2}{[1,2,3]}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\ ],[4,5]]}{[100,2,3]}\\
    \tab = [[100,2,3],[\ ],[4,5]]$\\

@@

$\pg{phead_1 \circ phead_2}{[[1,2,3],[\ ],[4,5]]}{100}\\
    \tab = (f, a) \Leftarrow \pg{phead_1}{[[1,2,3],[\ ],[4,5]}{dummy};\\
        \qtab (b, c) \Leftarrow \pg{phead_2}{a}{100};\\
        \qtab (d, e) \Leftarrow \pg{phead_1}{f}{b};\\
            \qtab \tab (d, c)\\
    \tab = (f, a) \Leftarrow ([dummy,[\ ],[4,5]], [1,2,3]);\\
        \qtab (b, c) \Leftarrow \pg{phead_2}{[1,2,3]}{100};\\
        \qtab (d, e) \Leftarrow \pg{phead_1}{f}{b};\\
            \qtab \tab (d, c)\\
    \tab = (f, a) \Leftarrow ([dummy,[\ ],[4,5]], [1,2,3]);\\
        \qtab (b, c) \Leftarrow ([100,2,3], 1);\\
        \qtab (d, e) \Leftarrow \pg{phead_1}{[dummy,[\ ],[4,5]]}{[100,2,3]};\\
            \qtab \tab (d, c)\\
    \tab = (f, a) \Leftarrow ([dummy,[\ ],[4,5]], [1,2,3]);\\
        \qtab (b, c) \Leftarrow ([100,2,3], 1);\\
        \qtab (d, e) \Leftarrow ([[100,2,3],[\ ],[4,5]], dummy);\\
            \qtab \tab (d, c)\\
    \tab = ([[100,2,3],[\ ],[4,5]], 1)$\\

In the above evaluation, we do not clearly define what $dummy$ is. As we mentioned, $dummy$ is a special value which makes the first $\pginline{phead_1}$ valid. In this case, if we care about the type of $dummy$, it should be a list of integers.

\textcolor{red}{
    \begin{itemize}
    \item In pg approach we always need to supply source and view.
    \item If we would like to evaluate get, we do not have any view.
    \item For this, we require the programmer to write a function to produce view from source. (This can be the same with get??)
    \end{itemize}
}

%\begin{itemize}
%\item Because we combine put and get, we need to supply view for evaluation of get.
%\item introduce dummy?
%\item automatic analysis for domain \& range?
%\item 
%\item partial evaluation of put(s(D), dummy(S)).
%\item \textcolor{red}{Not good? Basically put (BX) is often a recursion on the source. If source is static, partial evaluation will work well. On the other hand, view is small data. So partial evaluation might not work well.}
%\item \textcolor{red}{lazy evaluation is also a possibility?}
%\end{itemize}

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, we realize that there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given BX program is a left-associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable.

@@

To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we will use some following utility functions:

>>>>

\newcommand{\sif}[3]{\text{if } #1 \text{ then } #2 \text{ else } #3}
\newcommand{\product}[2]{#1 \times #2}
\newcommand{\tuple}[2]{(#1 :: #2)}
\newcommand{\rearrs}[3]{RearrS \ #1 \ #2 \ #3}
\newcommand{\rearrv}[3]{RearrV \ #1 \ #2 \ #3}
\newcommand{\casebx}[4]{Case \ #1 \ #2 \ #3 \ #4}

@@

MinBiGUL, our target language in this paper, is a subset of BiGUL [?] which is a simple yet powerful putback-based bidirectional language. BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are following functions:\\
    $\tab \getbxinline{bx} = fun \ s \to v \\
    \tab \putbxinline{bx} = fun \ s \ v \to v$

BiGUL is well-behaved since two functions $\putbxinline{bx}$ and $\getbxinline{bx}$ satify the round-trip laws as follows:\\
    $\tab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\text{GETPUT}]\\
    \tab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\text{PUTGET}]$

The GETPUT law means that if there is no change to the view, there should be no change to the source. The PUTGET law means that we can recover the modified view by applying the forward transformation to the updated source.

MinBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which are satify two above laws. In addition, we restrict adaptive cases of BiGUL on minBiGUL. Then $put$ and $get$ satify one more law, PUTPUT, like the following:\\
    $\tab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\text{PUTPUT}]$

The PUTPUT law means that a source update should overwrite the effect of previous source updates.

Due to the satisfaction of three laws, GETPUT, PUTGET and PUTPUT, minBiGUL is very well-behaved.

In the remaining of this section, we will present about the syntax and semantics of minBiGUL through formal definitions and examples.

@@

The syntax of minBiGUL is briefly written as follows:

$bx := Skip \ h \\ 
    \qtab | \ Replace \\ 
    \qtab | \ Prod \ bx_1 \ bx_2 \\ 
    \qtab | \ RearrS \ f_1 \ f_2 \ bx \\
    \qtab | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtab | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \\ 
    \qtab | \ Compose \ bx_1 \ bx_2$

A minBiGUL program may be either a skip of a function or a replacement or a product of two minBiGUL programs or a source / view rearrangement or a case combinatator without adaptive cases or a composition of some minBiGUL programs.

For source / view rearrangement, BiGUL use just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need give two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

To help make demonstration more direct, we provide the following alternatives representation:

    $\tab Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2$\\
    $\tab Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$

In general, $\circ$ has a higher priority than $\times$. Since only consider a product of two minBiGUL programs is considered, it is unnecessary to define the associativity precedence of $\times$. The one of $\circ$ may be left or right, however we do not declare the default.

In the next part, we will introduce the semantics of minBiGUL.

\subsection{Semantics}

\begin{definition}
    $\putbx{bx}{s}{v}$

@@

    $\putbx{\product{bx_1}{bx_2}}{(s_1, s_2)}{(v_1, v_2)} =\\
        \tab ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}))$

@@

    $\putbx{bx_1 \circ bx_2}{s}{v} = \putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}$
\end{definition}

\begin{definition}
    $\getbx{bx}{s}$

@@

    $\getbx{\product{bx_1}{bx_2}}{(s_1,s_2)} = \\
        \tab ((\getbx{bx_1}{s_1}),(\getbx{bx_2}{s_2}))$

@@

    $\getbx{bx_1 \circ bx_2}{s} = \getbx{bx_2}{(\getbx{bx_1}{s})}$
\end{definition}

In defintions 1 and 2, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tab \equiv \text{if }E_1 = true \text{ then}\\ 
    \qtab \tab X_1;\\ 
    \qtab \tab \text{if } E_2 = true \text{ then } S \text{ else assert } false;\\
    \qtab \text{else}\\ 
    \qtab \tab X_2;\\ 
    \qtab \tab \text{if } E_2 = false \text{ then } S \text{ else assert } false;$

The if-then-else-fi statement is useful to describe many functions related to $Case$ in this paper.

Next, let's take a look at some examples to better understand about minBiGUL. We start with quite obivious things as follows:\\
$\putbx{Skip \ (\lambda x.(x*x))}{10}{100} = 10$\\
$\putbx{Skip \ (\lambda \_.())}{1}{()} = 1$\\ 
$\getbx{Skip \ (\lambda x.(x*x))}{10} = 100$\\
$\getbx{Skip \ (\lambda \_.())}{1} = ()$\\
$\putbx{Replace}{1}{100} = 100$\\
$\getbx{Replace}{1} = 1$\\

Now, let's consider the definition of $phead$ in BiGUL [?]:

$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where:}\\
    \qtab f_1 = \lambda (s::ss).(s,ss)\\
    \qtab f_2 = \lambda (s,ss).(s::ss)\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where:}\\
        \qtab \qtab g_1 = \lambda v.(v,())\\
        \qtab \qtab g_2 = \lambda (v,()).v\\
        \qtab \qtab bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$\\

Then, we can evaluate $\putbxinline{phead}$ and $\getbxinline{phead}$ based on their definitions:

$\putbx{phead}{[1,2,3]}{100}\\
\tab = \putbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}{100}\\
\tab = f_2 \ (\putbx{bx_s}{(f_1 \ [1,2,3])}{100})\\
\tab = f_2 \ (\putbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}{100})\\
\tab = f_2 \ (\putbx{bx_v}{(1,[2,3])}{(g_1 \ 100}))\\
\tab = f_2 \ (\putbx{\product{Replace}{(Skip \ (\lambda \_.()))}}{(1,[2,3])}{(100,())})\\
\tab = f_2 \ ((\putbx{Replace}{1}{100}), (\putbx{(Skip \ (\lambda \_.()))}{[2,3]}{()}))\\
\tab = f_2 \ (100, [2,3])\\
\tab = [100,2,3]$\\

$\getbx{phead}{[1,2,3]}\\
\tab = \getbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}\\
\tab = \getbx{bx_s}{(f_1 \ [1,2,3])}\\
\tab = \getbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}\\
\tab = g_2 \ (\getbx{bx_v}{(1,[2,3])})\\
\tab = g_2 \ (\getbx{\product{Replace}{(Skip \ (\lambda \_.()))}}{(1,[2,3])})\\
\tab = g_2 \ ((\getbx{Replace}{1}), (\getbx{(Skip \ (\lambda \_.()))}{[2,3]}))\\
\tab = g_2 \ (1,())\\
\tab = 1$\\

    \tab $\putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$\\
    \tab $\getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$\\

$\getbx{phead_1 \circ phead_2}{[[1,2,3],[\,],[4,5]]}\\
    \tab = \getbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\,],[4,5]]})}\\
    \tab = \getbx{phead_2}{[1,2,3]}\\
    \tab = 1$\\

$\putbx{phead_1 \circ phead_2}{[[1,2,3],[\,],[4,5]]}{100}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{(\putbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\,],[4,5]]})}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{(\putbx{phead_2}{[1,2,3]}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{[100,2,3]}\\
    \tab = [[100,2,3],[\,],[4,5]]$\\

One more example we would like to present is $replaceAll$ whose definition in BiGUL as follows:\\
$\tab replaceAll :: (Eq \ s, Show \ s) \Rightarrow BiGUL \ [s] \ s \\ 
\tab replaceAll = Case \ [\\
    \qtab \$(normal \ [\![\lambda s \ v \to length \ s == 1]\!][\![\lambda s \to length \ s == 1]\!]) \\ 
    \qtab \Longrightarrow \$(rearrS \ [\![\lambda [x] \to x]\!]) Replace,\\
    \qtab \$(normal \ [\![\lambda s \ v \to length \ s > 1]\!][\![\lambda s \to length \ s > 1]\!]) \\ 
    \qtab \Longrightarrow \$(rearrS \ [\![\lambda (x:xs) \to (x,xs)]\!])\$\\ 
    \qtab \qtab \$(rearrV \ [\![\lambda v \to (v,v)]\!])\$\\ 
    \qtab \qtab Replace \ `prod\text{'} \ replaceAll\\
\tab ]$\\

The following expresses $replaceAll$ in minBiGUL:\\
$\tab replaceAll = Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \ \text{where:}\\ 
    \qtab cond_{sv} = \lambda s.\lambda v.length \ s == 1\\
    \qtab cond_{s} = \lambda s.length \ s == 1\\ 
    \qtab bx_1 = RearrS \ (\lambda [x] \to x) \ (\lambda x \to [x]) \ Replace\\ 
    \qtab bx_2 = RearrS \ (\lambda (x::xs) \to (x,xs)) \ (\lambda (x,xs) \to (x::xs)) \ bx_s \ \text{where:}\\ 
        \qtab \tab bx_s = RearrV \ (\lambda v \to (v,v)) \ (\lambda (v,v) \to v) \ bx_v \ \text{where:}\\
            \qtab \qtab bx_v = \product{Replace}{replaceAll}$

We can also evaluate $\putbxinline{replaceAll}$ and $\getbxinline{replaceAll}$ similarly to the previous examples if we pay attention to conditions $cond_sv$ and $cond_s$ as well as the if-then-esle-fi structure. $\putbxinline{replaceAll}$ uses the view to update all elements in the source list, while $\getbxinline{replaceAll}$ returns the an element of the source list with same elements. For example, $\putbx{replaceAll}{[1,2,3]}{100} = [100,100,100]$ and $\getbx{replaceAll}{[1,1,1]} = 1$.

@@

In minBiGUL or even BiGUL, when evaluating a composition of many programs, $get$s are re-evaluated so many times since no intermediate state is stored during the evaluation. This is a kind of information loss. One question is whether it is possible to calculate such programs without losing information. And the answer is yes. It comes from the the idea of reversible computation where all information during the evaluation need to be kept. In minBiGUL, we can do that by tupling $put$ and $get$. A pair of a source and a view is accepted as the input of a function named $pg$ to produce a new pair that contains the actual result of a corresponding minBiGUL program.

@@

$\pg{\product{bx_1}{bx_2}}{(s_1,s_2)}{(v_1,v_2)} \\
    \tab = ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}), (\getbx{bx_1}{s_1}), (\getbx{bx_2}{s_2})) \\
    \tab = (s_1, v_1) \Leftarrow \pg{bx_1}{s_1}{v_1}; \\
        \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{s_2}{v_2}; \\
        \qtab \tab ((s_1,s_2), (v_1,v_2))$

@@

Constructions of $pg$ for the replacement, the product and the source / view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.\\

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling $put$ and $get$ occurs, there will be 4 combinations of these conditions. This means that two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satified. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others which is used in the construction of $\pginline{Case}$ above. \\

@@

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. It is possible if we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid. Due to GETPUT law, we can use $get \ s$ to update the source $s$ in the $put$ direction. Then it is feasible to construct a dummy from the source. The last equality changes $dummy$ by an application $construct\_dummy \ s$, and also substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which is the evaluated result of $\putbx{bx_1}{s}{construct\_dummy \ s}$. This transformation is possible due to the PUTPUT law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

@@

On the other hand, we can evaluate $\pginline{phead}$ due to the construction of $pg$:
$\pg{phead}{[1,2,3]}{100}\\
    \tab = \pg{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}{100}\\
    \tab = (s, v) \Leftarrow \pg{bx_s}{f_1 \ [1,2,3]}{100};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \pg{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}{100};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \pg{bx_v}{(1,[2,3])}{g_1 \ 100};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \pg{\product{Replace}{(Skip \ (\lambda \_.()))}}{(1,[2,3])}{((100,()))};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \{\\
            \qtab \qtab (s_1, v_1) \Leftarrow \pg{Replace}{1}{100};\\
            \qtab \qtab (s_2, v_2) \Leftarrow \pg{(Skip \ (\lambda \_.()))}{[2,3]}{()};\\
            \qtab \qtab ((s_1,s_2), (v_1,v_2))\\
        \qtab \tab \};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow \{\\
            \qtab \qtab (s_1, v_1) \Leftarrow (100, 1);\\
            \qtab \qtab (s_2, v_2) \Leftarrow ([2,3], ());\\
            \qtab \qtab ((s_1,s_2), (v_1,v_2))\\
        \qtab \tab \};\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow \{\\
        \qtab \tab (s, v) \Leftarrow ((100,[2,3]), (1,()))\\
        \qtab \tab (s, g_2 \ v)\\
        \qtab \};\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = (s, v) \Leftarrow ((100, [2,3]), 1);\\
        \qtab \tab (f_2 \ s, v)\\
    \tab = ([100,2,3], 1)$\\

@@

$\pg{phead_1 \circ phead_2}{[[1,2,3],[\,],[4,5]]}{100}\\
    \tab = (s_1, v_1) \Leftarrow \pg{phead_1}{[[1,2,3],[\,],[4,5]}{dummy};\\
        \qtab (s_2, v_2) \Leftarrow \pg{phead_2}{v_1}{100};\\
        \qtab (s_3, v_3) \Leftarrow \pg{phead_1}{s_1}{s_2};\\
            \qtab \tab (s_3, v_2)\\
    \tab = (s_1, v_1) \Leftarrow ([dummy,[\,],[4,5]], [1,2,3]);\\
        \qtab (s_2, v_2) \Leftarrow \pg{phead_2}{[1,2,3]}{100};\\
        \qtab (s_3, v_3) \Leftarrow \pg{phead_1}{s_1}{s_2};\\
            \qtab \tab (s_3, v_2)\\
    \tab = (s_1, v_1) \Leftarrow ([dummy,[\,],[4,5]], [1,2,3]);\\
        \qtab (s_2, v_2) \Leftarrow ([100,2,3], 1);\\
        \qtab (s_3, v_3) \Leftarrow \pg{phead_1}{[dummy,[\,],[4,5]]}{[100,2,3]};\\
            \qtab \tab (s_3, v_2)\\
    \tab = (s_1, v_1) \Leftarrow ([dummy,[\,],[4,5]], [1,2,3]);\\
        \qtab (s_2, v_2) \Leftarrow ([100,2,3], 1);\\
        \qtab (s_3, v_3) \Leftarrow ([[100,2,3],[\,],[4,5]], dummy);\\
            \qtab \tab (s_3, v_2)\\
    \tab = ([[100,2,3],[\,],[4,5]], 1)$\\

In the above evaluation, we do not clearly contruct what $dummy$ is. As we mentioned, $dummy$ is a special value which makes the first $\pginline{phead_1}$ valid. In this case, if we care about the type of $dummy$, it should be a list of integers. Thus, we can choose $dummy$ is an arbitrary list which has that type, for instance, an empty list or the head element of the source.

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, we realize that there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given program is a left-associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable.

@@

To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we provide some following utility functions:

$\tab fst = \lambda (x_1,x_2). x_1\\
 \tab snd = \lambda (x_1,x_2). x_2\\
 \tab con = \lambda ks_1. \lambda ks_2. \lambda x. ((ks_1 \ x),(ks_2 \ x))$\\
